PLANETESIMAL ACCRETION AT SHORT ORBITAL PERIODS

### 2023-07-13 main.tex
Fix broken citations, change paths to data in python script

Planetesimal accretion is a key phase in the terrestrial planet growth
process, bridging the gap from kilometer-sized bodies up to roughly
moon-sized objects known as planetary embryos. In the earliest stages
of the planet formation process, beginning from $\mu$m sizes, aerodynamic forces dominate the
growth and evolution of the solids and statistical models
\cite{johansen14, birnstiel16} are appropriate to describe how these
numerous, small bodies coagulate.

Due to the internal pressure support
of the gas disk, the gas itself orbits at sub-Keplerian speed and
exerts a headwind on any solids large enough to decouple from the gas
\cite{weidenschilling77}. Around a meter in size, this headwind
is maximally effective at sapping away orbital angular momentum, and planet-building material can fall onto the central star on 
catastrophically short timescales \cite{weidenschilling77, nakagawa86}. Additionally, laboratory experiments suggest that 
collisions between mm- to cm- sized solids tend to result in bounces or destruction, rather than continued growth
\cite{blum93, colwell03, beitz11}.

For these reasons, a number of mechanisms to radially concentrate solids in a planet-forming disk have been proposed to 
facilitate fast growth from mm to km sizes \cite{johansen07, lyra08, bai10} in order to surmount these barriers. Interestingly, 
formation models for the short-period multiplanet systems revealed by Kepler \cite{fabrycky14} also seem to require enhanced 
concentrations of planet-building material to reproduce the observed architectures \cite{raymond07, hansen12}.

Regardless of how the mm- to km-sized growth barriers are surmounted, gravity begins to dominate and aerodynamic gas drag 
plays an increasingly unimportant role beyond this size. During this phase, collision cross sections are enhanced as gravitational 
focusing \cite{safronov69} acts to bend the trajectories of bodies undergoing close encounters. Because gravitational focusing 
becomes more effective as bodies grow larger, a period of runaway growth occurs \cite{wetherill89, kokubo96, barnes09} and a 
power law spectrum of masses develop. Eventually, the largest bodies (known as oligarchs) dynamically heat the remaining 
planetesimals, severely limiting further growth \cite{kokubo98}. The final outcome of this phase is a bimodal population of 
dynamically cold oligarchs, surrounded by dynamically hot, difficult to accrete residual planetesimals. Lines of evidence suggest 
that the asteroid belt \cite{bottke05, morbidelli09}, Kuiper belt \cite{duncan89, levison08, sheppard10} and the Oort cloud  \cite{levison11} are largely composed of the leftovers of this stage of planet formation.

Tidal interactions between protoplanets and the gaseous disk keep eccentricities and inclinations low until the gas disk dissipates, typically over the course of a few Myr \cite{mamajek09}. \textbf{On a timescale of roughly $10^{5}$ yr, gravitational perturbations trigger an instability which involves large scale oscillations of the eccentricity and inclinations of bodies as they strongly interact through secular resonances \cite{chambers98}.}
As a consequence of the instability, the oligarchics are no longer
on isolated, stable orbits and coalesce to form Earth-sized planets
through a series of extremely energetic giant 
impacts \cite{kokubo02, raymond05, raymond06}.

Due to the relative ease of modeling the early dust coagulation phases
and the final giant impact phase, these steps in the terrestrial
planet formation process have received most of the attention in the
literature. The planetesimal accretion phase, which we will focus on
in this paper, is more difficult and expensive to model because there are too
many particles to directly track with traditional N-body codes, while
the gravitational interactions between the few massive bodies produced by the runaway growth phase \cite{ida93, kokubo95, kokubo98} render statistical methods inappropriate. Due to computational expense, N-body simulations of
planetesimal accretion are usually modeled in a narrow
ring \cite{kokubo96, kokubo98}, and the results and timescales are then scaled to suit whatever relevant orbital period is being 
studied. N-body simulations of terrestrial planet assembly typically begin with a series of neatly spaced oligarchs, whose mass 
varies smoothly with orbital period. As we will show in this paper, oligarchic growth does not scale to arbitrarily short
orbital periods.

Given that Systems of Tightly-packed Inner Planets (STIPs) appear to
be a common outcome of planet formation \cite{lantham11, lissauer11, rowe14}, understanding exactly how
solids accumulate at short orbital periods is crucial. Although
gas-disk driven migration of the planets themselves is often invoked
to explain the observed architectures \cite{izidoro17, izidoro21}, we will focus on an in-situ
model in this paper. That is, once the planetesimals themselves form,
they largely stay in place, and any subsequent large-scale movement of
the solids are the result of mutual gravitational interactions. The
focus of this work will be to understand how the outcome of the
planetesimal accretion process scales with orbital period by using a
high-powered N-body code to directly follow the growth and evolution
of the planetesimals across a wide range of orbital periods (1 to 100
days). In doing so, we will assess whether the typical initial conditions (fully formed, evenly spaced protoplanets, e.g. \cite{raymond06}) used in studies of terrestrial planet formation are actually appropriate for understanding STIPs.

@@

We begin our analysis by considering a disk of equal mass planetesimals
with radius $r_{pl}$, mass $m_{pl}$ and surface density
$\Sigma_{pl}$. The collision rate in the vicinity of an orbit defined
by Keplerian frequency $\Omega$ can be written as $n \Gamma v$, where
$n = \Sigma_{pl} \Omega / 2 m_{pl} v$ (where we have assumed that the scale height of the planetesimal disk goes as $2v/\Omega$). $\Gamma$ describes the effective
collision cross section and $v$ is the typical encounter velocity
between planetesimals.
For a swarm of planetesimals on randomly oriented orbits, $v$ is typically
taken to be the rms velocity, $\left< v^{2} \right>^{1/2}$, which can be related to the eccentricity and inclination distribution $(e, i)$ in the following way \cite{lissauer93}:
\begin{equation}\label{eq:ecc_vel}
	\left< v^{2} \right>^{1/2} = \left( \frac{5}{4} \left< e^{2} \right>^{1/2} + \left< i^{2} \right>^{1/2}  \right) v_{k}.
\end{equation}

@@

Assuming that every collision results in a perfect merger, the growth rate of a planetesimal is given by
\begin{equation}\label{eq:growth}
	\frac{1}{M}\frac{dM}{dt} = \frac{\Sigma \Omega}{2 m_{pl}} \Gamma.
\end{equation}
\noindent In the case where the collision cross section depends only
on the physical size of the planetesimals, the growth scales sub-linearly
with mass and the mass distribution is expected to evolve in an
``orderly'' fashion, in which mass ratios between bodies tend toward unity. However, bodies larger than $\sim 100$ km in size are expected to exert a significant gravitational force on each other during encounters and the collision cross section depends on both the size of the bodies and their encounter velocities. In this case, 
\begin{equation}\label{eq:gravfocus}
	\Gamma = \pi r_{pl}^2 \left( 1 + v_{esc}^2 / v^2 \right)
\end{equation}
\cite{safronov69}, where $v_{esc}$ is the escape velocity from the two bodies at the point of contact.

In the limit that $v_{esc} \gg v$, it can be shown that $dM/dt \propto
M^{4/3}$, which implies a runaway scenario where growth
accelerates with mass. This mode of growth was confirmed with N-body
simulations by \cite{kokubo96} and appears necessary to construct
protoplanets within the lifetime of a protoplanetary disk \cite{lissauer87}, although one should note that pebble accretion \cite{lambrechts12, lambrechts14, bitsch15} is a viable alternative scenario. Due to the
velocity dependence of the gravitational focusing effect, it is not clear how ubiquitous this mode of growth is. In particular, encounter velocities at short orbital periods will be rather large (because $v \sim v_{k}$) and the $v_{esc} \gg v$ condition may not always be satisfied. The effect that a dynamically hot disk has on runway growth will be examined in detail in section \ref{sec:narrow}.

An important feature is missing from the model described above, which
limits its applicability at late times. Gravitational stirring, which converts
Keplerian shear into random motion, raises the typical encounter velocity
between planetesimals over time \cite{weidenschilling89, ida90} and diminishes the effectiveness of gravitational
focusing. As the mass spectrum of the system evolves away from uniformity, these velocity differences become even more
pronounced. As the system evolves, it tends toward a state of
energy equipartition where $v \sim m^{1/2}$. For a system of equal mass bodies in which encounters are driven by random 
motions rather than Keplerian shear (dispersion dominated), the timescale for gravitational stirring is described by the two-body 
relaxation time \cite{ida93}
\begin{equation}\label{eq:relax}
	t_{relax} = \frac{v^3}{4 \pi n G^2 {m_{pl}}^2 \ln \Lambda},
\end{equation}
where $\ln \Lambda$ is the Coulomb logarithm,
typically taken to be $\approx 3$ for a planetesimal disk \cite{ida90, stewart00}. Despite
the fact that the behavior of gravitational stirring is well-described
by a two-body formula, \cite{ida93} found that the stirring in a planetesimal disk is actually driven by close encounters, which 
requires a three-body formalism. As we will show in section \ref{sec:narrow}, gravitational stirring effectively shuts off when the 
Hill sphere of a body becomes comparable to its physical size. In this case, close encounters tend to result in collisions, and the 
main pathway for energy exchange between planetesimals and growing protoplanets is unable to operate.

\cite{kokubo98} showed that the runaway growth process described above is actually self-limiting. As the runaway
bodies develop, they become increasingly effective at dynamically heating the
remaining planetesimals, which diminishes the
gravitational focusing cross sections and throttles the growth rate. Around the time
that the mass of the runaway bodies exceeds the mass of the planetesimals
by a factor of $\sim 50-100$
\cite{ida93} a phase of less vigorous ``oligarchic'' 
growth commences, in which the largest bodies continue to 
accrete planetesimals at similar rates, independent of mass.

@@

\begin{equation}\label{eq:iso_mass1}
	M_{iso} = 4 \pi a^{2} \tilde{b} \left(\frac{M_{iso}}{3 M_{*}} \right)^{1/3} \Sigma_{pl}.
\end{equation}
\noindent Solving for $M_{iso}$ gives
\begin{equation}\label{eq:iso}
	M_{iso} = \left[ \frac{\left( 2 \pi a^2 \Sigma_{pl} \tilde{b} \right)^3}{3 M_{*}} \right]^{1/2}.
\end{equation}
\noindent For bodies on circular, non-inclined orbits, $\tilde{b} = 2 \sqrt{3}$ is the smallest orbital separation that produces a non-negative
Jacobi energy and permits a close encounter \cite{naka88}. This value of $\tilde{b}$ is typically used to calculate the final isolation mass of a protoplanet because oligarchic growth tends to maintain near-circular orbits for the growing protoplanets.

@@

In figure \ref{fig:timescales}, we show the ratio between the relaxation
and collision timescale for a population of equal-mass planetesimals
as a function of orbital period. Here, the encounter velocity is
described by equation \ref{eq:ecc_vel}. For
simplicity, we assume that $\left< e^2 \right>^{1/2} = 2\left< i^2
\right>^{1/2}$ \cite{ida93a} and that the eccentricity dispersion is
constant with orbital period. The coloring indicates the ratio between
$t_{relax}$ and $t_{coll}$. The dashed lines denote the eccentricity at which the
random encounter velocity (calculated according to equation \ref{eq:ecc_vel}) is
equal to the mutual escape velocity of the bodies. This is shown for planetesimals
with an internal density of 3 g cm$^{-3}$ ranging from 10 to 200 km in size.

A physically realistic value of $e_{h}$ for a population of planetesimals is going to depend on the structure
of the gaseous disk (which we will address in section \ref{sec:fulldisk}). However, the eccentricity
dispersion will typically increase over time until $\left< v^{2} \right>^{1/2} = v_{esc}$ and this is
often used to construct the initial conditions (e.g. \cite{barnes09}). Therefore, the curves in this figure
should be interpreted as upper limits.

@@

We also introduce a second dimensionless quantity, which relates the physical size of the bodies to the velocity state of the system
\begin{equation}\label{eq:beta}
	\beta = \frac{r_{pl}}{r_{g}}.
\end{equation}
where $r_{g} = G m_{pl} / v^{2}$ is the gravitational radius of a body (see eq 4.1 of \cite{ida90}). Encounters between bodies inside of a distance of $r_{g}$ result in significant deflections of their trajectories. It should be noted that the gravitational focusing enhancement factor $v^{2}/v_{esc}^{2}$ is equal to 1 for $\beta = 1$. In the case where $r_{g}$ is smaller than the size of a planetesimal, the gravitational focusing enhancement factor will be between 0 and 1 and the collision cross section is mostly set by the geometric value. For very large values of $r_{g}$ ($\beta \ll 1$), the effective collision cross section is almost entirely set by gravitational scattering.

@@

We use the tree-based N-body code {\sc ChaNGa} to model the gravitational and collisional evolution of a swarm of planetesimals. {\sc ChaNGa} is written using the {\sc CHARM++} parallel programming language and has been shown to perform well on up to half a million processors \cite{menon15} and can follow the evolution of gravitationally interacting collections of up to billions of particles. Using a modified Barnes-Hut tree with hexadecapole expansions of the moments to approximate forces, {\sc ChaNGa} integrates the equations of motion using a kick-drift-kick leapfrog scheme. For all of the simulations presented in this paper, we use a node opening criteria of $\theta_{BH}$ = 0.7. Additional information about the code is available in \cite{jetley08,menon15}.

Using the neighbor-finding algorithm in {\sc ChaNGa}, originally
designed for SPH calculations, we have recently implemented a solid
body collision module in the code.  This work is largely based on the
solid-body collision implementation in {\sc PKDGRAV}, which is
described in \cite{richardson94} and \cite{richardson00}. To
summarize, imminent collisions are detected during the ``drift'' phase
by extrapolating positions of bodies forward in time, using the
velocity calculated at the opening ``kick''. For each body, any
neighboring particles which fall within a search ball of radius
$2 \Delta T v + 2 r_{pl}$, where $\Delta T$ is the current timestep size
for the particle and $v$ is magnitude of its heliocentric velocity,
 are tested for an imminent collision. In the case that
a collision is detected, the particles are merged into a single larger
body, which is given the center of mass position and velocity of the
two parents. Resolving a collision can produce another imminent
collision, so collisions are handled one-by-one and another full
collision check is run after the previous event is resolved. For a
more detailed description of the collision module in {\sc ChaNGa}, see
\cite{wallace19}.
Particles are advanced on individual timesteps chosen as a
power of two of a base timestep.  The timestep for an individual
particle is based on an estimate of the gravitational dynamical time
determined by the minimum of $\sqrt{d_{node}^3/(G(M_{node} + m_{pl}))}$
across all nodes in the tree that are accepted by the Barnes-Hut
opening criterion.  Here $d_{node}$ is the distance from the
planetesimal to the center of mass of the tree node and $M_{node}$ is
the total mass of the tree node.  For nearby particles $M_{node}$ is
replaced with the mass of the nearby particle.

@@

To answer these questions, we run a series of simulations in which a
narrow annulus of planetesimals orbits a star. The values of $\alpha$
and $\beta$ are varied individually. 4000 planetesimals with
individual masses of $8.37 \times 10^{-5} M_{\earth}$ are placed with semimajor
axes drawn from a uniform distribution between 0.95 and 1.05 AU about a 1 $M_{\odot}$
star. In total, the disk contains $\sim$ 0.33 $M_{\earth}$ of material. The argument of perihelion, $\omega$, longitude of ascending node,
$\Omega$, and mean anomaly, M, for each body is drawn from a uniform
distribution $\in [0, 2 \pi)$. The inclinations and eccentricities are drawn
from a Rayleigh distribution with
$\left< i^{2} \right> = 1/2 \left< e^{2} \right>$ \cite{ida93a}.

In the ``fiducial'' case, we give the bodies a bulk density of 3 g
cm$^{-3}$ (for a radius of 341 km), and $\left< e^{2} \right>^{1/2} = 4 e_{h}$, which corresponds to $\alpha = 3.6 \times 10^{-2}$ and $\beta = 3.4 \times 10^{-3}$. These parameters are chosen to match the initial conditions of \cite{kokubo98}, which gave rise to oligarchic growth. 
To vary the value of $\alpha$, we alter the bulk density of the particles. In the high-$\alpha$ case, the bulk density is reduced by 
a factor of $\sim$ 7100, which produces $\alpha = 1$. This corresponds to a bulk density of $4.2 \times 10^{-4}$ g cm$^{-3}$ and a radius of 6,500 km. 
Although this is most certainly unphysical, the purpose of this modification is to have a planetesimal completely fills its Hill sphere so that no strong 
gravitational scattering should occur. To vary $\beta$, the eccentricity dispersion is increased. For the high-$
\beta$ case, $\left< e^{2} \right>^{1/2}$ is increased to $1500 e_{h}$, which corresponds to $\beta = 15,000$. This is the largest value of $\beta$ that permits a particle at 1 AU to still have an apocenter and pericenter distance that lies within the boundaries of the disk. This choice of $\beta$ places the system firmly in the $v > v_{esc}$ regime, while still allowing growth to occur in a reasonable number of timesteps.

@@

Additionally, these results suggest that the value of $\alpha$, which is a function of only the initial conditions, controls the qualitative outcome of accretion. 
Across most of a planet-forming disk, $\alpha$ is small, and frequent gravitational encounters between the growing bodies will 
facilitate oligarchic growth. In the dispersion-dominated regime, close encounters drive the stirring between planetesimals and 
embryos \cite{weidenschilling89, ida90}. When $\alpha \ll 1$, the \textbf{planetesimal fills only a small portion of its Hill sphere} and the majority of close encounters result in viscous stirring, rather than accretion. In the opposite regime, we 
observe that runaway growth still occurs, but nearly all of the planetesimals are consumed by the forming protoplanets, rather 
than being scattered onto higher eccentricity orbits, where they would otherwise remain as a remnant of the early stages of 
planet formation \cite{kokubo98, kokubo00}.

@@

Motivated by the qualitative dependence of accretion on $\alpha$,
we next investigate whether this highly efficient, non-oligarchic
growth should be expected to operate near the innermost regions of a typical planet-forming
disk. Given that N-body simulations of short-period terrestrial planet formation 
typically begin with a chain of neatly-spaced, isolation mass 
(see \cite{kokubo00} eq. 20) protoplanets, it is pertinent to determine 
whether the high $\alpha$ growth mode we revealed
in the previous section invalidates this choice of initial conditions.

>>>>

Planetesimal accretion is a key phase in the terrestrial planet growth
process, bridging the gap from kilometer-sized bodies up to roughly
moon-sized objects known as planetary embryos. In the earliest stages
of the planet formation process, beginning from $\mu$m sizes, aerodynamic forces dominate the
growth and evolution of the solids and statistical models
\citep{johansen14, birnstiel16} are appropriate to describe how these
numerous, small bodies coagulate.

Due to the internal pressure support
of the gas disk, the gas itself orbits at sub-Keplerian speed and
exerts a headwind on any solids large enough to decouple from the gas
\citep{weidenschilling77}. Around a meter in size, this headwind
is maximally effective at sapping away orbital angular momentum, and planet-building material can fall onto the central star on 
catastrophically short timescales \citep{weidenschilling77, nakagawa86}. Additionally, laboratory experiments suggest that 
collisions between mm- to cm- sized solids tend to result in bounces or destruction, rather than continued growth
\citep{blum93, colwell03, beitz11}.

For these reasons, a number of mechanisms to radially concentrate solids in a planet-forming disk have been proposed to 
facilitate fast growth from mm to km sizes \citep{johansen07, lyra08, bai10} in order to surmount these barriers. Interestingly, 
formation models for the short-period multiplanet systems revealed by Kepler \citep{fabrycky14} also seem to require enhanced 
concentrations of planet-building material to reproduce the observed architectures \citep{raymond07, hansen12}.

Regardless of how the mm- to km-sized growth barriers are surmounted, gravity begins to dominate and aerodynamic gas drag 
plays an increasingly unimportant role beyond this size. During this phase, collision cross sections are enhanced as gravitational 
focusing \citep{safronov69} acts to bend the trajectories of bodies undergoing close encounters. Because gravitational focusing 
becomes more effective as bodies grow larger, a period of runaway growth occurs \citep{wetherill89, kokubo96, barnes09} and a 
power law spectrum of masses develop. Eventually, the largest bodies (known as oligarchs) dynamically heat the remaining 
planetesimals, severely limiting further growth \citep{kokubo98}. The final outcome of this phase is a bimodal population of 
dynamically cold oligarchs, surrounded by dynamically hot, difficult to accrete residual planetesimals. Lines of evidence suggest 
that the asteroid belt \citep{bottke05, morbidelli09}, Kuiper belt \citep{duncan89, levison08, sheppard10} and the Oort cloud  \citep{levison11} are largely composed of the leftovers of this stage of planet formation.

Tidal interactions between protoplanets and the gaseous disk keep eccentricities and inclinations low until the gas disk dissipates, typically over the course of a few Myr \citep{mamajek09}. \textbf{On a timescale of roughly $10^{5}$ yr, gravitational perturbations trigger an instability which involves large scale oscillations of the eccentricity and inclinations of bodies as they strongly interact through secular resonances \citep{chambers98}.}
As a consequence of the instability, the oligarchics are no longer
on isolated, stable orbits and coalesce to form Earth-sized planets
through a series of extremely energetic giant 
impacts \citep{kokubo02, raymond05, raymond06}.

Due to the relative ease of modeling the early dust coagulation phases
and the final giant impact phase, these steps in the terrestrial
planet formation process have received most of the attention in the
literature. The planetesimal accretion phase, which we will focus on
in this paper, is more difficult and expensive to model because there are too
many particles to directly track with traditional N-body codes, while
the gravitational interactions between the few massive bodies produced by the runaway growth phase \citep{ida93, kokubo95, kokubo98} render statistical methods inappropriate. Due to computational expense, N-body simulations of
planetesimal accretion are usually modeled in a narrow
ring \citep{kokubo96, kokubo98}, and the results and timescales are then scaled to suit whatever relevant orbital period is being 
studied. N-body simulations of terrestrial planet assembly typically begin with a series of neatly spaced oligarchs, whose mass 
varies smoothly with orbital period. As we will show in this paper, oligarchic growth does not scale to arbitrarily short
orbital periods.

Given that Systems of Tightly-packed Inner Planets (STIPs) appear to
be a common outcome of planet formation \citep{lantham11, lissauer11, rowe14}, understanding exactly how
solids accumulate at short orbital periods is crucial. Although
gas-disk driven migration of the planets themselves is often invoked
to explain the observed architectures \citep{izidoro17, izidoro21}, we will focus on an in-situ
model in this paper. That is, once the planetesimals themselves form,
they largely stay in place, and any subsequent large-scale movement of
the solids are the result of mutual gravitational interactions. The
focus of this work will be to understand how the outcome of the
planetesimal accretion process scales with orbital period by using a
high-powered N-body code to directly follow the growth and evolution
of the planetesimals across a wide range of orbital periods (1 to 100
days). In doing so, we will assess whether the typical initial conditions (fully formed, evenly spaced protoplanets, e.g. \citet{raymond06}) used in studies of terrestrial planet formation are actually appropriate for understanding STIPs.

@@

We begin our analysis by considering a disk of equal mass planetesimals
with radius $r_{pl}$, mass $m_{pl}$ and surface density
$\Sigma_{pl}$. The collision rate in the vicinity of an orbit defined
by Keplerian frequency $\Omega$ can be written as $n \Gamma v$, where
$n = \Sigma_{pl} \Omega / 2 m_{pl} v$ (where we have assumed that the scale height of the planetesimal disk goes as $2v/\Omega$). $\Gamma$ describes the effective
collision cross section and $v$ is the typical encounter velocity
between planetesimals.
For a swarm of planetesimals on randomly oriented orbits, $v$ is typically
taken to be the rms velocity, $\left< v^{2} \right>^{1/2}$, which can be related to the eccentricity and inclination distribution $(e, i)$ in the following way \citep{lissauer93}:
\begin{equation}\label{eq:ecc_vel}
	\left< v^{2} \right>^{1/2} = \left( \frac{5}{4} \left< e^{2} \right>^{1/2} + \left< i^{2} \right>^{1/2}  \right) v_{k}.
\end{equation}

@@

Assuming that every collision results in a perfect merger, the growth rate of a planetesimal is given by
\begin{equation}\label{eq:growth}
	\frac{1}{M}\frac{dM}{dt} = \frac{\Sigma \Omega}{2 m_{pl}} \Gamma.
\end{equation}
\noindent In the case where the collision cross section depends only
on the physical size of the planetesimals, the growth scales sub-linearly
with mass and the mass distribution is expected to evolve in an
``orderly'' fashion, in which mass ratios between bodies tend toward unity. However, bodies larger than $\sim 100$ km in size are expected to exert a significant gravitational force on each other during encounters and the collision cross section depends on both the size of the bodies and their encounter velocities. In this case, 
\begin{equation}\label{eq:gravfocus}
	\Gamma = \pi r_{pl}^2 \left( 1 + v_{esc}^2 / v^2 \right)
\end{equation}
\citep{safronov69}, where $v_{esc}$ is the escape velocity from the two bodies at the point of contact.

In the limit that $v_{esc} \gg v$, it can be shown that $dM/dt \propto
M^{4/3}$, which implies a runaway scenario where growth
accelerates with mass. This mode of growth was confirmed with N-body
simulations by \citet{kokubo96} and appears necessary to construct
protoplanets within the lifetime of a protoplanetary disk \citep{lissauer87}, although one should note that pebble accretion \citep{lambrechts12, lambrechts14, bitsch15} is a viable alternative scenario. Due to the
velocity dependence of the gravitational focusing effect, it is not clear how ubiquitous this mode of growth is. In particular, encounter velocities at short orbital periods will be rather large (because $v \sim v_{k}$) and the $v_{esc} \gg v$ condition may not always be satisfied. The effect that a dynamically hot disk has on runway growth will be examined in detail in section \ref{sec:narrow}.

An important feature is missing from the model described above, which
limits its applicability at late times. Gravitational stirring, which converts
Keplerian shear into random motion, raises the typical encounter velocity
between planetesimals over time \citep{weidenschilling89, ida90} and diminishes the effectiveness of gravitational
focusing. As the mass spectrum of the system evolves away from uniformity, these velocity differences become even more
pronounced. As the system evolves, it tends toward a state of
energy equipartition where $v \sim m^{1/2}$. For a system of equal mass bodies in which encounters are driven by random 
motions rather than Keplerian shear (dispersion dominated), the timescale for gravitational stirring is described by the two-body 
relaxation time \citep{ida93}
\begin{equation}\label{eq:relax}
	t_{relax} = \frac{v^3}{4 \pi n G^2 {m_{pl}}^2 \ln \Lambda},
\end{equation}
where $\ln \Lambda$ is the Coulomb logarithm,
typically taken to be $\approx 3$ for a planetesimal disk \citep{ida90, stewart00}. Despite
the fact that the behavior of gravitational stirring is well-described
by a two-body formula, \citep{ida93} found that the stirring in a planetesimal disk is actually driven by close encounters, which 
requires a three-body formalism. As we will show in section \ref{sec:narrow}, gravitational stirring effectively shuts off when the 
Hill sphere of a body becomes comparable to its physical size. In this case, close encounters tend to result in collisions, and the 
main pathway for energy exchange between planetesimals and growing protoplanets is unable to operate.

\citet{kokubo98} showed that the runaway growth process described above is actually self-limiting. As the runaway
bodies develop, they become increasingly effective at dynamically heating the
remaining planetesimals, which diminishes the
gravitational focusing cross sections and throttles the growth rate. Around the time
that the mass of the runaway bodies exceeds the mass of the planetesimals
by a factor of $\sim 50-100$
\citep{ida93} a phase of less vigorous ``oligarchic'' 
growth commences, in which the largest bodies continue to 
accrete planetesimals at similar rates, independent of mass.

@@

\begin{equation}\label{eq:iso_mass1}
	M_{iso} = 4 \pi a^{2} \tilde{b} \left(\frac{M_{iso}}{3 M_{*}} \right)^{1/3} \Sigma_{pl}.
\end{equation}
\noindent Solving for $M_{iso}$ gives
\begin{equation}\label{eq:iso}
	M_{iso} = \left[ \frac{\left( 2 \pi a^2 \Sigma_{pl} \tilde{b} \right)^3}{3 M_{*}} \right]^{1/2}.
\end{equation}
\noindent For bodies on circular, non-inclined orbits, $\tilde{b} = 2 \sqrt{3}$ is the smallest orbital separation that produces a non-negative
Jacobi energy and permits a close encounter \citep{naka88}. This value of $\tilde{b}$ is typically used to calculate the final isolation mass of a protoplanet because oligarchic growth tends to maintain near-circular orbits for the growing protoplanets.

@@

In figure \ref{fig:timescales}, we show the ratio between the relaxation
and collision timescale for a population of equal-mass planetesimals
as a function of orbital period. Here, the encounter velocity is
described by equation \ref{eq:ecc_vel}. For
simplicity, we assume that $\left< e^2 \right>^{1/2} = 2\left< i^2
\right>^{1/2}$ \citep{ida93a} and that the eccentricity dispersion is
constant with orbital period. The coloring indicates the ratio between
$t_{relax}$ and $t_{coll}$. The dashed lines denote the eccentricity at which the
random encounter velocity (calculated according to equation \ref{eq:ecc_vel}) is
equal to the mutual escape velocity of the bodies. This is shown for planetesimals
with an internal density of 3 g cm$^{-3}$ ranging from 10 to 200 km in size.

A physically realistic value of $e_{h}$ for a population of planetesimals is going to depend on the structure
of the gaseous disk (which we will address in section \ref{sec:fulldisk}). However, the eccentricity
dispersion will typically increase over time until $\left< v^{2} \right>^{1/2} = v_{esc}$ and this is
often used to construct the initial conditions (e.g. \citet{barnes09}). Therefore, the curves in this figure
should be interpreted as upper limits.

@@

We also introduce a second dimensionless quantity, which relates the physical size of the bodies to the velocity state of the system
\begin{equation}\label{eq:beta}
	\beta = \frac{r_{pl}}{r_{g}}.
\end{equation}
where $r_{g} = G m_{pl} / v^{2}$ is the gravitational radius of a body (see eq 4.1 of \citet{ida90}). Encounters between bodies inside of a distance of $r_{g}$ result in significant deflections of their trajectories. It should be noted that the gravitational focusing enhancement factor $v^{2}/v_{esc}^{2}$ is equal to 1 for $\beta = 1$. In the case where $r_{g}$ is smaller than the size of a planetesimal, the gravitational focusing enhancement factor will be between 0 and 1 and the collision cross section is mostly set by the geometric value. For very large values of $r_{g}$ ($\beta \ll 1$), the effective collision cross section is almost entirely set by gravitational scattering.

@@

We use the tree-based N-body code {\sc ChaNGa} to model the gravitational and collisional evolution of a swarm of planetesimals. {\sc ChaNGa} is written using the {\sc CHARM++} parallel programming language and has been shown to perform well on up to half a million processors \citep{menon15} and can follow the evolution of gravitationally interacting collections of up to billions of particles. Using a modified Barnes-Hut tree with hexadecapole expansions of the moments to approximate forces, {\sc ChaNGa} integrates the equations of motion using a kick-drift-kick leapfrog scheme. For all of the simulations presented in this paper, we use a node opening criteria of $\theta_{BH}$ = 0.7. Additional information about the code is available in \citep{jetley08,menon15}.

Using the neighbor-finding algorithm in {\sc ChaNGa}, originally
designed for SPH calculations, we have recently implemented a solid
body collision module in the code.  This work is largely based on the
solid-body collision implementation in {\sc PKDGRAV}, which is
described in \citet{richardson94} and \citet{richardson00}. To
summarize, imminent collisions are detected during the ``drift'' phase
by extrapolating positions of bodies forward in time, using the
velocity calculated at the opening ``kick''. For each body, any
neighboring particles which fall within a search ball of radius
$2 \Delta T v + 2 r_{pl}$, where $\Delta T$ is the current timestep size
for the particle and $v$ is magnitude of its heliocentric velocity,
 are tested for an imminent collision. In the case that
a collision is detected, the particles are merged into a single larger
body, which is given the center of mass position and velocity of the
two parents. Resolving a collision can produce another imminent
collision, so collisions are handled one-by-one and another full
collision check is run after the previous event is resolved. For a
more detailed description of the collision module in {\sc ChaNGa}, see
\citep{wallace19}.
Particles are advanced on individual timesteps chosen as a
power of two of a base timestep.  The timestep for an individual
particle is based on an estimate of the gravitational dynamical time
determined by the minimum of $\sqrt{d_{node}^3/(G(M_{node} + m_{pl}))}$
across all nodes in the tree that are accepted by the Barnes-Hut
opening criterion.  Here $d_{node}$ is the distance from the
planetesimal to the center of mass of the tree node and $M_{node}$ is
the total mass of the tree node.  For nearby particles $M_{node}$ is
replaced with the mass of the nearby particle.

@@

To answer these questions, we run a series of simulations in which a
narrow annulus of planetesimals orbits a star. The values of $\alpha$
and $\beta$ are varied individually. 4000 planetesimals with
individual masses of $8.37 \times 10^{-5} M_{\earth}$ are placed with semimajor
axes drawn from a uniform distribution between 0.95 and 1.05 AU about a 1 $M_{\odot}$
star. In total, the disk contains $\sim$ 0.33 $M_{\earth}$ of material. The argument of perihelion, $\omega$, longitude of ascending node,
$\Omega$, and mean anomaly, M, for each body is drawn from a uniform
distribution $\in [0, 2 \pi)$. The inclinations and eccentricities are drawn
from a Rayleigh distribution with
$\left< i^{2} \right> = 1/2 \left< e^{2} \right>$ \citep{ida93a}.

In the ``fiducial'' case, we give the bodies a bulk density of 3 g
cm$^{-3}$ (for a radius of 341 km), and $\left< e^{2} \right>^{1/2} = 4 e_{h}$, which corresponds to $\alpha = 3.6 \times 10^{-2}$ and $\beta = 3.4 \times 10^{-3}$. These parameters are chosen to match the initial conditions of \citet{kokubo98}, which gave rise to oligarchic growth. 
To vary the value of $\alpha$, we alter the bulk density of the particles. In the high-$\alpha$ case, the bulk density is reduced by 
a factor of $\sim$ 7100, which produces $\alpha = 1$. This corresponds to a bulk density of $4.2 \times 10^{-4}$ g cm$^{-3}$ and a radius of 6,500 km. 
Although this is most certainly unphysical, the purpose of this modification is to have a planetesimal completely fills its Hill sphere so that no strong 
gravitational scattering should occur. To vary $\beta$, the eccentricity dispersion is increased. For the high-$
\beta$ case, $\left< e^{2} \right>^{1/2}$ is increased to $1500 e_{h}$, which corresponds to $\beta = 15,000$. This is the largest value of $\beta$ that permits a particle at 1 AU to still have an apocenter and pericenter distance that lies within the boundaries of the disk. This choice of $\beta$ places the system firmly in the $v > v_{esc}$ regime, while still allowing growth to occur in a reasonable number of timesteps.

@@

Additionally, these results suggest that the value of $\alpha$, which is a function of only the initial conditions, controls the qualitative outcome of accretion. 
Across most of a planet-forming disk, $\alpha$ is small, and frequent gravitational encounters between the growing bodies will 
facilitate oligarchic growth. In the dispersion-dominated regime, close encounters drive the stirring between planetesimals and 
embryos \citep{weidenschilling89, ida90}. When $\alpha \ll 1$, the \textbf{planetesimal fills only a small portion of its Hill sphere} and the majority of close encounters result in viscous stirring, rather than accretion. In the opposite regime, we 
observe that runaway growth still occurs, but nearly all of the planetesimals are consumed by the forming protoplanets, rather 
than being scattered onto higher eccentricity orbits, where they would otherwise remain as a remnant of the early stages of 
planet formation \citep{kokubo98, kokubo00}.

@@

Motivated by the qualitative dependence of accretion on $\alpha$,
we next investigate whether this highly efficient, non-oligarchic
growth should be expected to operate near the innermost regions of a typical planet-forming
disk. Given that N-body simulations of short-period terrestrial planet formation 
typically begin with a chain of neatly-spaced, isolation mass 
(see \citet{kokubo00} eq. 20) protoplanets, it is pertinent to determine 
whether the high $\alpha$ growth mode we revealed
in the previous section invalidates this choice of initial conditions.
